### 2025-09-30 misc/expo/idempotent.tex
posted tutorial 3

\maketitle{Idempotent theorems and matrices}{}{Marcel K. Goh}{\sl Department of Mathematics and Statistics,
McGill University}

@@

\edef\propopencriterion{\the\thmcount}
\proclaim Proposition \advthm. A set $A$ is open if and only if for every $x\in A$, there is an open
set $U_x$ with $x\in U_x$ and $U_x\subseteq A$.

@@

A set $A$ is said to be a {\it neighbourhood} of a point $p$ if $p$ belongs to the interior of $A$.
The space $S$ is called {\it Hausdorff} if for all points $x,y\in S$ with $x\ne y$, there is some
open neighbourhood $U$ of $X$ and open neighbourhood $V$ of $y$ such that $U\cap V=\emptyset$.
A set $A$ is {\it compact} if every open cover of $A$ has a finite subcover. If every point of $S$
has some compact neighbourhood, then $S$ is called {\it locally compact}.

If the singleton set $\{p\}$ is open, then $p$ is called an {\it isolated point} of $S$, and if
$p$ is isolated for all points $p\in S$, then $S$ is a {\it discrete space}.

@@

\edef\propnbhdnet{\the\thmcount}
\proclaim Proposition \advthm.
Let $x$ be a point in a topological space $S$ and let $N(x)$ be the set of all open neighbourhoods
of $x$. Define an order on $N(x)$ by declaring $U\preceq V$ if and only if $U\supset V$. Then the order
$\preceq$ is directed, and $x$ is a limit point of any
net $(x_U)_{U\in N(x)}$ satisfying $x_U\in U$ for all $U\in N(x)$.

\proof The set $N(x)$ is clearly directed, since for every $U,V\in N(x)$,
$U \supset U\cap V$ and $V\supset U\cap V$. Now let $(x_U)_{U\in N(x)}$ be any net with
$x_U\in U$ for all $U\in N(x)$. Pick any open set $V$ containing $x$. This set $V$ is a member of
$N(x)$, and for any $U\in N(x)$ with $U\succeq V$, we must have $x_U\in U \subseteq V$.\slug

@@

\proclaim Proposition \advthm. A set $A$ is closed if and only if whenever a
net with $(x_i)_{i\in I}$ satisfying $x_i\in A$ for all $i\in I$ possesses a limit point $x$,
we must have $x\in A$.

@@

>>>>

\maketitle{Idempotent theorems and matrices}{}{Marcel K. Goh}{23 September 2025}

@@

\edef\propopencriterion{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. A set $A$ is open if and only if for every $x\in A$, there is an open
set $U_x$ with $x\in U_x$ and $U_x\subseteq A$.

@@

A set $A$ is said to be a {\it neighbourhood} of a point $p$ if $p$ belongs to the interior of $A$.
The space $S$ is called {\it Hausdorff} if for all points $x,y\in S$ with $x\ne y$, there is some
open neighbourhood $U$ of $X$ and open neighbourhood $V$ of $y$ such that $U\cap V=\emptyset$.

If the singleton set $\{p\}$ is open, then $p$ is called an {\it isolated point} of $S$, and if
$p$ is isolated for all points $p\in S$, then $S$ is a {\it discrete space}. It is easy to see
that every subset of a discrete space is open.

@@

\edef\propnbhdnet{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm.
Let $x$ be a point in a topological space $S$ and let $N(x)$ be the set of all open neighbourhoods
of $x$. Define an order on $N(x)$ by declaring $U\preceq V$ if and only if $U\supseteq V$. Then the order
$\preceq$ is directed, and $x$ is a limit point of any
net $(x_U)_{U\in N(x)}$ satisfying $x_U\in U$ for all $U\in N(x)$.

\proof The set $N(x)$ is clearly directed, since for every $U,V\in N(x)$,
$U \supseteq U\cap V$ and $V\supseteq U\cap V$. Now let $(x_U)_{U\in N(x)}$ be any net with
$x_U\in U$ for all $U\in N(x)$. Pick any open set $V$ containing $x$. This set $V$ is a member of
$N(x)$, and for any $U\in N(x)$ with $U\succeq V$, we must have $x_U\in U \subseteq V$.\slug

@@

\proclaim Proposition \advthm. A set $A$ is closed if and only if for every net
$(x_i)_{i\in I}$ satisfying $x_i\in A$ for all $i\in I$, every limit point $x$ of
$(x_i)_{i\in I}$
must be in $A$.

@@

\medskip\boldlabel Cluster points and subnets.
A point $x$ is a {\it cluster point} of the net $(x_i)_{i\in I}$ if for every open neighbourhood $U$
of $x$ and any $i\in I$, there is some $j\succeq i$ with $x_i\in U$. Next, consider two directed sets
$I$ and $J$. A map $\phi:J\to I$ is said to be {\it cofinal} if for every $i\in I$ there is $j\in J$
such that for all $j'\succeq j$, $\phi(j') \succeq i$. Now if $(x_i)_{i\in I}$ is a net (with elements
in some topological space $X$) and $\phi:B\to A$ is
cofinal, then the composition $x\circ \phi : B\to X$ is a net; we say that $x\circ \phi$ is
a {\it subnet} of the original net.
The following proposition relates subnets to cluster points.

\proclaim Proposition \advthm.
A point $x$ is a cluster point of the net $(x_i)_{i\in I}$ if and only if there is a subnet
of $(x_i)_{i\in I}$ converging to $x$.

\proof Suppose there is a subnet of $(x_i)_{i\in I}$ converging to $x$. That is, there is a directed
set $J$ and a cofinal map $\phi : J\to I$ such that, setting $y_j = x_{\phi(j)}$ for all $j\in J$,
the subnet $(y_j)_{j\in J}$ of $(x_i)_{i\in I}$ converges to $x$. Let $U$ be an open neighbourhood
of $x$ and let $i\in I$. There is some $j_0\in J$ such that $j\succeq j_0$ implies that $y_j\in U$.
Since $\phi$ is cofinal, there is $j_1\in J$ such that for all $j\succeq j_1$, $\phi(j)\succeq i$.
Now since $J$ is directed, there is some $j'\in J$ with $j'\succeq j_0$ and $j'\succeq j_1$.
For this choice of $j'$ we have $\phi(j')\succeq i$ and $x_{\phi(j')} = y_{j'} \in U$. Hence $x$
is a cluster point of $(x_i)_{i\in I}$.

Conversely, suppose that $x$ is a cluster point of the net $(x_i)_{i\in I}$. Let $N(x)$
be the set of all neighbourhoods of $x$ as in Proposition~{\propnbhdnet}. We define
$$J = \bigl\{ (U, i) : U\in N(x), x_i\in U\bigr\}$$
and impose the preorder $(U,i) \preceq (U', i')$ if and only if $U\supseteq U'$ and $i\preceq i'$.
For any $(U_0, i_0)$ and $(U_1, i_1)$ in $J$, we can pick $i'\in I$ such that $i\succeq i_0$
and $i\succeq i_1$. Since $x$ is a cluster point, there is some $i^*\ge i$ such that
$x_{i^*}\in U_0\cap U_1$, and we see that $(U_0, i_0) \preceq (U_0\cap U_1, i^*)$
and $(U_1, i_1)\preceq (U_0 \cap U_i, i^*)$. This shows that $J$ is a directed set. We define
the map $\phi : J\to I$ by sending $(U,i)\mapsto i$. This is certainly cofinal, by our choice of preorder
on $J$. To show that $\bigl(x_{\phi(j)}\bigr)_{j\in J}$ converges to $x$, let $V$ be any neighbourhood
of $x$. Since $x$ is a cluster point, there is some $i\in I$ with $x_i\in V$; consider the
element $j = (V, i)\in J$. For any $j' = (V', i') \in J$ with $(V', i')\succeq (V,i)$, we have
$V' \subseteq V$ and $x_{i'}\in V'$, so $x_{\phi(j')} = x_{i'}\in V$.\slug

\medskip\boldlabel Compactness.
A set $A$ is {\it compact} if every open cover of $A$ has a finite subcover. If every point of $S$
has some compact neighbourhood, then $S$ is called {\it locally compact}.
